Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-20 paper/paper.tex
Rewrite the first part of the intro

\section{Introduction}
% Depending on its application, we care about certain metatheoretic
% properties about a type system. As a programming language, we may care
% about type soundness, which states that a well-typed never gets stuck
% during evaluation.
When a dependently type system is used as a program logic where terms
encode proofs, we want our type system to be logically consistent,
meaning that the empty type is not inhabited.

>>>>

\section{Introduction}
In this paper, we present a short and mechanized consistency proof by
logical relation for \lang{}, a Martin-Löf style dependent type theory
with large elimination, a full predicative universe hierarchy, an
intensional identity type and a boolean base type, both of which
equipped with dependent elimination forms. % The mechanization is
% extremely short and involves less than 1000 lines of manually written
% Coq code.

Our goal is to show the fundamental technique required to carry out proof by logical
relation in a dependently typed setting where logical relation can no
longer be defined as a simple recursive function over the structure of
types. Compared to the many mechanized proofs by logical relation for
predicative dependent type
theory~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards},
whose proof developments range from 20,000 to 400,000 lines of code,
our Coq development is the most lightweight at less than 1,000 lines.
The big difference in the size of the developments does not
necessarily imply that our proof technique leads to a more concise
proof due to the differing expressiveness of the languages and the
different metatheoretic results being established. For
example, \citet{anand2014towards} mechanizes the metatheory of
Nuprl~\citep{constable1986implementing}, which, on top of all the
features that \lang{} supports, includes W-types and partial types and
requires a PER semantics to model its extensional typed
equality. The object language from \citet{decagda} does not support
identity types and only has one predicative universe, but additionally
supports $\Sigma$ types. However, for a researcher who wishes to learn
the underlying techniques for mechanized logical relation for
dependent type theory, a small development like ours is much easier to
navigate and understand. Furthermore, we show how our simple
consistency proof can be easily extended to show the existence of
$\beta\eta$-normal form for well-typed open \emph{and} closed terms,
giving us a metatheoretic result almost as strong as the one from
\citet{decagda} at the moderate cost of around 500 lines of extra code.

We present our definitions and proofs in set theory notation and
terminology rather than in Coq syntax to make the material accessible
to readers with general mathematical background. However, we make the
typeset definitions as close as possible to our mechanization so the
definitions and the proofs can be easily encoded in a proof
assistant.
We find the close correspondence
between our typeset presentation and the Coq development advantageous
for a couple of reasons.

% Depending on its application, we care about certain metatheoretic
% properties about a type system. As a programming language, we may care
% about type soundness, which states that a well-typed never gets stuck
% during evaluation.
When a dependently type system is used as a program logic where terms
encode proofs, we want our type system to be logically consistent,
meaning that the empty type is not inhabited.
